\section{Reading Comprehension  Questions}

\noindent
{\em From Section~\ref{section:dProofs}}

\begin{requ}
Because it was (perhaps incorrectly) assumed that you have heard the term {\em proof} before, 
it was never formally defined in the chapter. Let's make sure you don't go any further without having a good definition.
So, what is a {\em proof}? Feel free to look up the definition (online or in a dictionary) if you need to.
\begin{requanswer}
Answers will vary, but it should say something like 
``An argument using logic and/or math to demonstrate or show the nature of a conclusion,''
or 
``the process or an instance of establishing the validity of a statement especially by derivation 
from other statements in accordance with principles of reasoning'' (the latter is from Merriam-Webster). 
\end{requanswer}
\end{requ}

\begin{requ}
Let's say someone correctly proves statement $A$. 
Does that mean $A$ is a true statement, that you are just pretty sure that it is true, 
or that it may or may not be true
based on whether or not you understand the argument being made in the proof? 
Explain your answer.
\begin{requanswer}
If someone correctly proves statement $A$, that means it is a true statement, regardless of whether or not
you understand the proof.  That's because, by definition, a proof establishes the validity of a statement.
\end{requanswer}
\end{requ}

\begin{requ}
True or false: Every even number is not odd. Explain your answer.
\begin{requanswer}
True. An even integer is one of the form $2k$, where $k$ is an integer. 
An odd integer is one of the form $2k + 1$ where $k$ is an integer. 
Thus, even numbers are divisible by 2 whereas odd numbers are not. 
A number cannot both be divisible by 2 and not divisible by 2 at the same time.
\end{requanswer}
\end{requ}

\begin{requ}
If $b$ is divisible by $a$, is it always the case that $a$ is divisible by $b$? Explain using an example.
\begin{requanswer}
No. For instance, 6 is divisible by 2, but 2 is not divisible by 6. 
\end{requanswer}
\end{requ}

\begin{requ}
Can a number be both {\em composite} and {\em prime}? Explain.
\begin{requanswer}
No.
A number cannot be both composite and prime because by definition,
a composite integer is a positive integer $c>1$ that is {\em not} prime.
Thus, every integer greater than $1$ is either prime or composite, but never both.
\end{requanswer}
\end{requ}

\begin{requ}
Which are {\em prime}? 
Which of the following numbers are {\em composite}?
Which are neither? Which are both?\\
1, 3, 4, 6, 38, 27, 97, 150, 173, 999983, 999985
\begin{requanswer}
3, 97, 173, and 999983 are prime.
4, 6, 27, 38, 150, and 999985 are composite.
1 is neither. It is impossible to be both.
\end{requanswer}
\end{requ}


\begin{requ}
Compute $6!$ and $7!$. Did you compute $7!$ the easy way or the hard way?
\begin{requanswer}
$6!=6\cdot 5\cdot 4\cdot 3\cdot 2\cdot 1 = 720$.
$7!=7\cdot 6! = 7*720 =5040,$ computed the easy way
be recognizing I just needed to multiply the previous answer by 7.
\end{requanswer}
\end{requ}


\begin{requ}
For what values of $n$ is $n!$ prime?
\begin{requanswer}
Clearly $2!=2$ is prime.
We will show that this is the only case in which $n!$ is prime. 
$1!=1$, which is not prime by definition.
$3!=2\cdot 3$, which is clearly not prime.
If $n>3$, $n!$ is divisible by $3!=3\cdot 2$, so it not prime.
\end{requanswer}
\end{requ}

\medskip

\noindent
{\em From Section~\ref{section:implication}}

%% This one is blurry and doesn't belong here anyway.
%\begin{requ}
%Is ``$x\geq 0$'' a proposition? Explain.
%\begin{requanswer}
%A proposition is which is either true or false, which means $x \geq 0$ is indeed a proposition because, 
%given a value of $x$, the statement is either true or false since either $x\geq 0$ (in which case it is true) 
%or $x<0$ (in which case it is false). 
%\end{requanswer}
%\end{requ}

\begin{requ}
If the proposition {\em $A$ implies $B$} is true, 
does that mean the proposition {\em $B$ implies $A$} is true?
Prove or give a counterexample. 
\begin{requanswer}
No. 
Take the example from the book, ``If you know Java, then you know a programming language'' (where $A$ is the proposition 
``you know Java'' and $B$ is the proposition ``you know a programming language.''
Since Java is a programming language, this proposition is true.
Then $B$ implies $A$ is the proposition ``if you know a programming language, then you know Java.'' But that is clearly not
true. You might know {\em C++}, which is a programming language, but not know Java.
\end{requanswer}
\end{requ}

\begin{requ}
True or false: If the inverse of an implication is true, then the implication is also true. Explain your answer.
\begin{requanswer}
False. From the previous answer, consider the implication ``if you know a programming language, then you know Java,'' 
and its inverse, ``if you do not know a programming language, then you do not know Java."
It shouldn't be too difficult to see that although the inverse is true, the implication is false.
\end{requanswer}
\end{requ}

\begin{requ}
True or false: If the inverse of an implication is true, then the converse of the implication is also true. Explain your answer.
\begin{requanswer}
This one is true since the inverse and converse of an implication are contrapositives of each other, and an implication 
and its contrapositive are equivalent.
\end{requanswer}
\end{requ}

\begin{requ}
What can you say about an implication and its contrapositive?
\begin{requanswer}
An implication and its contrapositive are equivalent.
Therefore, if one is true, then the other is true
(and if one is false, the other is false, of course). 
\end{requanswer}
\end{requ}

\medskip
 
\noindent
{\em From Sections~\ref{section:contradiction}
\begin{requ}
In your own words, explain the idea behind contradiction proofs. Include specifics like how one goes about
writing a proof by contradiction and why it is a valid proof technique.
(The goal of this question is to help you better understand the technique and to convince you that it is indeed
a valid technique, so put some thought into this one!)
\begin{requanswer}
Your answer will likely be different, and hopefully more detailed than the one provided here.
But you should say something about how you assume that what you want to prove is false, 
then use logic to arrive at a contradiction (that is, a statement that you know to be false).
Since you ``proved'' a false statement using correct logic, it must be that your premise (that is, 
the statement that you assumed was false) is incorrect. Since your premise was that the statement
you wanted to prove was false, then it must be that the statement is true (again, because you 
showed that if it is false, then you can prove something that is not true).
\end{requanswer}
\end{requ}

\begin{requ}
Give all of the permutations of the set $\{\mbox{cow},\mbox{chicken},\mbox{rabbit}\}$
\begin{requanswer}
They are
(cow, chicken, rabbit),
(cow, rabbit, chicken),
(chicken, cow, rabbit),
(chicken, rabbit, cow), 
(rabbit, cow, chicken), and
(rabbit, chicken, cow).
\end{requanswer}
\end{requ}

\begin{requ}\label{intIsRat}
True or false: Every integer is a rational number. Explain your answer.
\begin{requanswer}
True. Every integer $a$ can be written as $a=\frac{a}{1}$, where $a$ and $1$ are both integers and $1\neq 0$, so it is rational.
\end{requanswer}
\end{requ}

\begin{requ}
Prove that there is no smallest positive rational number.
(Hint: Use contradiction!)
\begin{requanswer}
Assume that $k$ is the smallest positive rational number.
Since $k$ is rational, we can write it as $p/q$ for integers
$p$ and $q\neq 0$. But clearly $k/2=p/(2q)$ is positive,
rational, and smaller than $k$, which contradicts our assumption
that $k$ was the smallest such number.
Therefore there is no smallest positive rational number.
\end{requanswer}
\end{requ}

\medskip
 
\noindent
{\em From Section~\ref{section:contraposition}

\begin{requ}
Explain why proof by contraposition is a valid proof technique.
\begin{requanswer}
Since an implication and its contrapositive are equivalent, if you prove the contrapositive of a implication is
true, then the implication must also be true. And that is exactly what proof by contraposition does.
\end{requanswer}
\end{requ}

\begin{requ}
Explain the difference between a proof by contradiction and a proof by contraposition, particularly
as it applies to proving statements of the form $p\rightarrow q$.
\begin{requanswer}
{\bf Contradiction:} Assume that $p$ and $\neg q$ are both true. Get a contradiction.
Conclude that if $p$ is true, $\neg q$ cannot be true, so $q$ must be true. Thus, $p\rightarrow q$ is true. 

\noindent
{\bf Contraposition: } Prove that $\neg q\rightarrow \neg p$, which is equivalent to $p\rightarrow q$.

\noindent
In both cases, you assume that $\neg q$ is true. Often the contradiction you get is that $\neg p$
is true (which is the goal in contraposition proof), 
so the majority of the proofs look the same. But they begin and end slightly differently.
\end{requanswer}
\end{requ}


\begin{requ}
True or false: Every irrational number is not an integer. Explain your answer.
\begin{requanswer}
True. From the Question~\ref{intIsRat}, we know that every integer is rational, so integers are not irrational. 
Therefore, if a number is irrational, it cannot be an integer.
Alternatively, you can recognize that this is essentially the contrapositive of the Question~\ref{intIsRat}, so it is also true.
\end{requanswer}
\end{requ}

\begin{requ}
Prove that if $x>0$ is irrational, then $\sqrt{x}$ is irrational using
\begin{enumerate}[(a)]
\item proof by contradiction.
\item proof by contraposition.
\end{enumerate}
\begin{requanswer}
(a) Assume that $x>0$ is irrational, but that $\sqrt{x}$ is rational.
Then $\sqrt{x}=p/q$ for some integers $p, q\neq 0$.
That means that $x=(\sqrt{x})^2=(p/q)^2=p^2/q^2$, which 
is clearly rational since $p^2$ and $q^2\neq 0$ are both rational.
But this contradicts our assumption that $x$ is irrational.
Therefore, $\sqrt{x}$ must be irrational.\\
(b) We will prove that if $\sqrt{x}$ is rational, then $x$ is rational,
which will imply our statement is true since this is the
contrapositive of our statement.
Assume $\sqrt{x}$ is rational. Then 
$\sqrt{x}=p/q$ for some integers $p, q\neq 0$.
Therefore, $x=(\sqrt{x})^2=(p/q)^2=p^2/q^2$
which is clearly rational since $p^2$ and $q^2\neq 0$ are both rational.
\end{requanswer}
\end{requ}



\medskip
 
\noindent
{\em From Section~\ref{section:other}}

\begin{requ}
True or false: Every rational number is an integer. Prove your answer.
\begin{requanswer}
False. For instance, $1.5=\frac{3}{2}$ is rational, but clearly not an integer.
\end{requanswer}
\end{requ}

\begin{requ}
Prove that an integer $n$ and $n^2$ have the same parity
(that is, they are both even or both odd).
\begin{requanswer}
We will use a proof by cases.
Case 1: If $n$ is even, then $n=2k$ for some integer $k$.
Then $n^2=(2k)^2=4k^2=2(2k^2)$, which is even since
$2k^2$ is an integer.
Case 2: If $n$ is odd, then $n=2k+1$ for some integer $k$.
Then $n^2=(2k+1)^2=4k^2+4k+1=2(2k^2+2k)+1$, which is odd since $2k^2+2k$ is an integer.
In both cases, the parity remains unchanged.
\end{requanswer}
\end{requ}

\medskip 

\noindent
{\em From Section~\ref{section:iff}

\begin{requ}
If you want to prove that {\em $A$ if and only if $B$} is true (where $A$ and $B$ are statements of some sort), 
can you just show that $A$ implies $B$?
If not, explain why that does not work and what you would have to do instead (or in addition).
\begin{requanswer}
No you would have to show the reverse as well. 
That is, if I want to prove $A$ if and only if $B$,
I would have to prove either the proposition ($A\rightarrow B$) or the contrapositive ($\neg B \rightarrow \neg A$) 
{\bf and} I would have to prove the inverse ($\neg A\rightarrow \neg B$) or the converse ($B\rightarrow A$). 
\end{requanswer}
\end{requ}

\begin{requ}
You want to prove that $p$ if and only if $q$ is true.
\begin{lenum}
\item
Is showing that $p$ implies $q$ and $\neg q$ implies $\neg p$
a valid technique?
Explain why or why not.
\item
Is showing that $\neg q\rightarrow\neg p$ and $q\rightarrow p$ 
a valid technique?
Explain why or why not.
\end{lenum}
\begin{requanswer}
(a) No. This is proving the forward direction twice by proving the implication and its contrapositive. 
One of these needs to be replaced with either the inverse or converse.
(b) Yes. This is proving the contrapositive and the converse.
\end{requanswer}
\end{requ}

\begin{requ}
Prove that an integer $n$ is even if and only if  $n^2$ is even.
\begin{requanswer}
For the forward direction, we assume that $n$ is even.
Then $n=2k$ for some integer $k$, 
so $n^2=(2k)^2=4k^2=2(2k^2)$, which is even since
$2k^2$ is an integer. Thus, $n$ even implies $n^2$ is even.\\
For the reverse direction, we will prove that
$n$ is not even implies $n^2$ is not even. In other words,
$n$ is odd implies that $n^2$ is odd.
Assume $n$ is odd. Then $n=2k+1$ for some integer $k$,
so $n^2=(2k+1)^2=4k^2+4k+1=2(2k^2+2k)+1$, which is odd since $2k^2+2k$ is an integer.
Thus, $n$ odd implies that $n^2$ is odd, which (as we have already
said, but we'll say it again anyway) is equivalent to  
$n^2$ is even implies $n$ is even.
\end{requanswer}
\end{requ}

\medskip 

\noindent
{\em From Sections~\ref{section:proofErrors}
\begin{requ}
Proof by counterexample is a valid proof technique. Proof by example is not. Explain the difference.
\begin{requanswer}
You use proof by counterexample to {\em disprove} statements. You only need to show 
one instance where the statement is not true to demonstrate that it is not always true, so proof by counterexample
is valid. On the other hand, proof by example is just demonstrating the truth of a statement given some specific values.
Just because it works for the given values, that does not prove that it works for all values, so it is not a valid proof technique.
\end{requanswer}
\end{requ}

\begin{requ}
If I want to prove some equation, should I write down the equation and work both sides until they are the same?
Explain why this is or is not a valid proof technique.
\begin{requanswer}
This is definitely {\em not} a valid proof technique! The problem is that when you write down an equation and start
working both sides of it, you are implicitly assuming that the equation you are starting with is true. But since you 
are trying to prove that it is true, you can't start your proof with the fact that it is true--that is circular reasoning. 
For instance, do you remember the supposed proof of  $-1=1$ that started by working both sides of the equation?
\end{requanswer}
\end{requ}


\begin{requ}
What is wrong with the following proof?
\begin{spf}
Assume $a=b$, where $a$ and $b$ are not zero.
Multiplying both sides by $a$, we get $a^2=ab$.
Subtracting $b^2$ from both sides, we get $a^2-b^2=ab-b^2$,
which is equivalent to $(a-b)(a+b)=(a-b)b$.
Dividing by $a-b$, we get $(a+b)=b$, which implies $2b=a$,
but since $a=b$, $2b=b$, and dividing by $b$, we finally conclude that
$2=1$.
\end{spf}
\begin{requanswer}
The step that divided both sides by $a-b$ was division by zero,
which is not allowed!
\end{requanswer}
\end{requ}
